Revisiting the Relationship between Short-Chain Fatty Acids, the Microbiota, and Colorectal Tumors

### 2019-03-28 submission/manuscript.Rmd
Add regression paragraph code

min_p_compare_tax <- w_wo_compare %>%
	filter(class=="adenoma" | class == "cancer") %>%
	filter(microbiome =="otu" | microbiome == "genus") %>%
  select(p.value) %>%
	min()

min_p_compare_metag <- w_wo_compare %>%
	filter(class=="adenoma" | class == "cancer") %>%
	filter(microbiome =="opf" | microbiome == "kegg") %>%
	select(p.value) %>%
	min()

@@

**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using Random Forest machine learning models [@Baxter2016; @Sze2018]. We repeated that analysis but added the concentration of the individual or pooled SCFAs as possible features to train a model. Models trained using SCFAs to classify individuals as having adenomas or carcinomas rather than normal colons had median areas under the receiver operator characteristic curve (AUROC) that were significantly greater than 0.5 (P~adenoma~<0.001 and P~carcinoma~<0.001); however, the AUROC values to detect the presence of adenomas or carcinomas were only `r random %>% filter(class=="adenoma") %>% pull(median)` and `r random %>% filter(class=="cancer") %>% pull(median)`, respectively (Figure 2A). When we trained the models with the SCFAs concentrations and operational taxonomic unit (OTU) or genus-level relative abundances the AUROC values were not significantly different from the models trained without the SCFA concentrations (P>`r min_p_compare_tax`; Figure 2A). We also trained models using a smaller dataset that generated shotgun metagenomic sequencing data from a subset of our cohort (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`) [@Hannigan2017]. We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories. Again, the performance of the models trained with the meteagenomic data did not improve when the SCFA concentrations were added as possible features when training the model (P>`r min_p_compare_metag`; Figure 2A). These data demonstrate that knowledge of the SCFA profile from a patient's fecal sample does not improve the ability to diagnose a colonic lesion.

```{r, regression}
```

**Knowledge of microbial community structure does not predict SCFA concentrations.** Regardless of a person's diagnosis, we next asked whether the fecal community structure was predictive of fecal SCFA concentrations. We trained Random Forest regression models using 16S rRNA gene sequence data binned into OTUs and genera to predict the concentration of the individual and pooled SCFAs. Regardless of the binning method or SCFA, the largest amount of variation that the model could explain was **0.XXXX** (Figure 2B). Next, we trained Random Forest regression models using metagenomic sequence data binned into OPFs and KEGG categories to predict the concentration of the individual and pooled SCFAs. Similar to the analysis using 16S rRNA gene sequence data, the metagenomic data was not predictive of SCFA concentration; the largest amount of variation that the models could explain was **0.XXXX** (Figure 2B). Because of the limited number of samples that we were able to generate metagenomic sequence data from, we used our 16S rRNA gene sequence data to impute metagenomes that were binned into metabolic pathways, enzyme commission numbers, or KEGG categories using PICRUSt. Again, SCFA concentrations could not be predicted based on the imputed metagenomic data; the largest amount of variation that the models could explain was **0.XXXX** (Figure 2B). The inability to model SCFA concentrations from microbiome data indicates that the knowledge of the abundance of organisms and genes is insufficient to predict SCFA concentrations.

**Conclusion.** Our data indicate that fecal SCFA concentrations are not associated with the presence of adenomas or carcinomas and that they provide weak predictive power to improve the ability to diagnose someone with one of these lesions. Furthermore, knowledge of the taxonomic and genetic structure of gut microbiota was not predictive of individual or pooled SCFA concentrations. These results complement existing literature that suggest that fiber consumption and the production of SCFAs are unable to prevent the risk of developing colonic tumors. It is important to note that our analysis concerned fecal SCFA concentrations and microbiome characterization and that observations along the mucosa near the site of lesions may provide a stronger association. Regardless, given the growing literature in this area, it is unlikely that SCFAs are the primary mechanism that limits tumorigenesis. This may be a cautionary result to temper enthusiasm for SCFAs as a biomarker of gut health more generally. Going forward it is critical to develop additional hypotheses for how the microbiome and host interact to drive tumorigenesis to better understand the disease process and identify biomarkers that will allow early detection of tumors.

@@

\includegraphics{../results/figures/scfa_abundance.pdf}

@@

@@

@@

>>>>

min_p_compare_tax <- w_wo_compare %>%
	filter(class=="adenoma" | class == "cancer") %>%
	filter(microbiome =="otu" | microbiome == "genus") %>%
  select(p.value) %>%
	min()
```

**Combining SCFA and microbiome data does not improve the ability to diagnose individual as having adenomas or carcinomas.** We previously found that binning 16S rRNA gene sequence data into operational taxonomic units based on 97% similarity or into genera enabled us to classify individuals as having adenomas or carcinomas using Random Forest machine learning models [@Baxter2016; @Sze2018]. We repeated that analysis but added the concentration of the SCFAs as possible features to train the models (Figure S1). Models trained using SCFAs to classify individuals as having adenomas or carcinomas rather than normal colons had median areas under the receiver operator characteristic curve (AUROC) that were significantly greater than 0.5 (P~adenoma~<0.001 and P~carcinoma~<0.001); however, the AUROC values to detect the presence of adenomas or carcinomas were only `r random %>% filter(class=="adenoma") %>% pull(median)` and `r random %>% filter(class=="cancer") %>% pull(median)`, respectively (Figure 2A). When we trained the models with the SCFAs concentrations and operational taxonomic unit (OTU) or genus-level relative abundances the AUROC values were not significantly different from the models trained without the SCFA concentrations (P>`r min_p_compare_tax`; Figure 2A). These data demonstrate that knowledge of the SCFA profile from a patient's fecal sample does not improve the ability to diagnose a colonic lesion.

```{r, regression}

regression_features <- c("genus", "otu", "kegg", "opf", "pc2pathways", "pc2ec")
scfa_types <- c("acetate", "propionate", "isobutyrate", "butyrate")

regression <- read_tsv("../data/rf/regression_data_pool.tsv",
												col_types=cols(class=col_character(),
													input=col_character(),
													.default=col_double()
												)
 											) %>%
							mutate(microbiome=str_replace(input, "fit_", ""),
										microbiome=str_replace(microbiome, "_fit", "")
								) %>%
							select(class, microbiome, test_Rsquared) %>%
							filter(class %in% scfa_types) %>%
							filter(microbiome %in% regression_features) %>%
							mutate(data_type = case_when(
					       microbiome %in% c("genus", "otu")  ~ "16S",
								 microbiome %in% c("kegg", "opf")  ~ "metagenomics",
								 microbiome %in% c("pc2ec", "pc2pathways")  ~ "picrust"
							 )) %>%
							group_by(class, microbiome, data_type) %>%
							summarize(median_Rsquared = median(test_Rsquared))

top_16S <- regression %>% filter(data_type == "16S") %>% arrange(desc(median_Rsquared)) %>% ungroup() %>% slice(1)

top_metagenomics <- regression %>% filter(data_type == "metagenomics") %>% arrange(desc(median_Rsquared)) %>% ungroup() %>% slice(1)

top_picrust <- regression %>% filter(data_type == "picrust") %>% arrange(desc(median_Rsquared)) %>% ungroup() %>% slice(1)

convert <- c(genus = "genus", otu = "OTU", kegg = "KEGG", opf = "OPF", pc2ec = "KEGG", pc2pathways = "pathway")

@@

**Knowledge of microbial community structure does not predict SCFA concentrations.** We next asked whether the fecal community structure was predictive of fecal SCFA concentrations, regardless of a person's diagnosis. We trained Random Forest regression models using 16S rRNA gene sequence data binned into OTUs and genera to predict the concentration of the SCFAs (Figure S2). The largest R^2^ between the observed SCFA concentrations and the modeled concentrations was `r top_16S$median_Rsquared` observed when using `r convert[top_16S$microbiome]` data to predict `r top_16S$class` concentrations (Figure 2B). We also used a smaller dataset of shotgun metagenomic sequencing data generated from a subset of our cohort (N~normal~=`r n_s %>% filter(dx == "normal") %>% pull(n)`, N~adenoma~=`r n_s %>% filter(dx == "adenoma") %>% pull(n)`, and N~cancer~=`r n_s %>% filter(dx == "cancer") %>% pull(n)`) [@Hannigan2017]. We binned genes extracted from the assembled metagenomes into operational protein families (OPFs) or KEGG categories and trained Random Forest regression models using metagenomic to predict the concentration of the SCFAs (Figure S2). Similar to the analysis using 16S rRNA gene sequence data, the metagenomic data was not predictive of SCFA concentration; the largest amount of variation that the models could explain was `r top_metagenomics$median_Rsquared`, which was observed when using `r convert[top_metagenomics$microbiome]` data to predict `r top_metagenomics$class` concentrations (Figure 2B). Because of the limited number of samples that we were able to generate metagenomic sequence data from, we used our 16S rRNA gene sequence data to impute metagenomes that were binned into metabolic pathways or KEGG categories using PICRUSt (Figure S2). SCFA concentrations could not be predicted based on the imputed metagenomic data. The largest amount of variation that the models could explain was `r top_picrust$median_Rsquared` observed when using `r convert[top_picrust$microbiome]` data to predict `r top_picrust$class` concentrations (Figure 2B). The inability to model SCFA concentrations from microbiome data indicates that the knowledge of the abundance of organisms and their genes was insufficient to predict SCFA concentrations.

**Conclusion.** Our data indicate that fecal SCFA concentrations are not associated with the presence of adenomas or carcinomas and that they provide weak predictive power to improve the ability to diagnose someone with one of these lesions. Furthermore, knowledge of the taxonomic and genetic structure of gut microbiota was not predictive SCFA concentrations. These results complement existing literature that suggest that fiber consumption and the production of SCFAs are unable to prevent the risk of developing colonic tumors. It is important to note that our analysis concerned fecal SCFA concentrations and microbiome characterization and that observations along the mucosa near the site of lesions may provide a stronger association. Regardless, given the growing literature in this area, it is unlikely that SCFAs are the primary mechanism that limits tumorigenesis. This may be a cautionary result to temper enthusiasm for SCFAs as a biomarker of gut health more generally. Going forward it is critical to develop additional hypotheses for how the microbiome and host interact to drive tumorigenesis to better understand the disease process and identify biomarkers that will allow early detection of tumors.

@@

\includegraphics{../results/figures/scfa_comparisons.pdf}

@@

\includegraphics{../results/figures/scfa_modelling.pdf}

@@

\includegraphics{../results/figures/classification_testing.pdf}

@@

\includegraphics{../results/figures/regression_testing.pdf}

### 2019-03-28 submission/manuscript.tex
Add regression paragraph code

\textbf{Combining SCFA and microbiome data does not improve the ability
to diagnose individual as having adenomas or carcinomas.} We previously
found that binning 16S rRNA gene sequence data into operational
taxonomic units based on 97\% similarity or into genera enabled us to
classify individuals as having adenomas or carcinomas using Random
Forest machine learning models (8, 16). We repeated that analysis but
added the concentration of the individual or pooled SCFAs as possible
features to train a model. Models trained using SCFAs to classify
individuals as having adenomas or carcinomas rather than normal colons
had median areas under the receiver operator characteristic curve
(AUROC) that were significantly greater than 0.5
(P\textsubscript{adenoma}\textless{}0.001 and
P\textsubscript{carcinoma}\textless{}0.001); however, the AUROC values
to detect the presence of adenomas or carcinomas were only 0.54 and
0.55, respectively (Figure 2A). When we trained the models with the
SCFAs concentrations and operational taxonomic unit (OTU) or genus-level
relative abundances the AUROC values were not significantly different
from the models trained without the SCFA concentrations
(P\textgreater{}0.21; Figure 2A). We also trained models using a smaller
dataset that generated shotgun metagenomic sequencing data from a subset
of our cohort (N\textsubscript{normal}=27, N\textsubscript{adenoma}=25,
and N\textsubscript{cancer}=26) (18). We binned genes extracted from the
assembled metagenomes into operational protein families (OPFs) or KEGG
categories. Again, the performance of the models trained with the
meteagenomic data did not improve when the SCFA concentrations were
added as possible features when training the model (P\textgreater{}0.24;
Figure 2A). These data demonstrate that knowledge of the SCFA profile
from a patient's fecal sample does not improve the ability to diagnose a
colonic lesion.

\textbf{Knowledge of microbial community structure does not predict SCFA
concentrations.} Regardless of a person's diagnosis, we next asked
whether the fecal community structure was predictive of fecal SCFA
concentrations. We trained Random Forest regression models using 16S
rRNA gene sequence data binned into OTUs and genera to predict the
concentration of the individual and pooled SCFAs. Regardless of the
binning method or SCFA, the largest amount of variation that the model
could explain was \textbf{0.XXXX} (Figure 2B). Next, we trained Random
Forest regression models using metagenomic sequence data binned into
OPFs and KEGG categories to predict the concentration of the individual
and pooled SCFAs. Similar to the analysis using 16S rRNA gene sequence
data, the metagenomic data was not predictive of SCFA concentration; the
largest amount of variation that the models could explain was
\textbf{0.XXXX} (Figure 2B). Because of the limited number of samples
that we were able to generate metagenomic sequence data from, we used
our 16S rRNA gene sequence data to impute metagenomes that were binned
into metabolic pathways, enzyme commission numbers, or KEGG categories
using PICRUSt. Again, SCFA concentrations could not be predicted based
on the imputed metagenomic data; the largest amount of variation that
the models could explain was \textbf{0.XXXX} (Figure 2B). The inability
to model SCFA concentrations from microbiome data indicates that the
knowledge of the abundance of organisms and genes is insufficient to
predict SCFA concentrations.

\textbf{Conclusion.} Our data indicate that fecal SCFA concentrations
are not associated with the presence of adenomas or carcinomas and that
they provide weak predictive power to improve the ability to diagnose
someone with one of these lesions. Furthermore, knowledge of the
taxonomic and genetic structure of gut microbiota was not predictive of
individual or pooled SCFA concentrations. These results complement
existing literature that suggest that fiber consumption and the
production of SCFAs are unable to prevent the risk of developing colonic
tumors. It is important to note that our analysis concerned fecal SCFA
concentrations and microbiome characterization and that observations
along the mucosa near the site of lesions may provide a stronger
association. Regardless, given the growing literature in this area, it
is unlikely that SCFAs are the primary mechanism that limits
tumorigenesis. This may be a cautionary result to temper enthusiasm for
SCFAs as a biomarker of gut health more generally. Going forward it is
critical to develop additional hypotheses for how the microbiome and
host interact to drive tumorigenesis to better understand the disease
process and identify biomarkers that will allow early detection of
tumors.

@@

>>>>

\textbf{Combining SCFA and microbiome data does not improve the ability
to diagnose individual as having adenomas or carcinomas.} We previously
found that binning 16S rRNA gene sequence data into operational
taxonomic units based on 97\% similarity or into genera enabled us to
classify individuals as having adenomas or carcinomas using Random
Forest machine learning models (8, 16). We repeated that analysis but
added the concentration of the SCFAs as possible features to train the
models (Figure S1). Models trained using SCFAs to classify individuals
as having adenomas or carcinomas rather than normal colons had median
areas under the receiver operator characteristic curve (AUROC) that were
significantly greater than 0.5 (P\textsubscript{adenoma}\textless{}0.001
and P\textsubscript{carcinoma}\textless{}0.001); however, the AUROC
values to detect the presence of adenomas or carcinomas were only 0.54
and 0.55, respectively (Figure 2A). When we trained the models with the
SCFAs concentrations and operational taxonomic unit (OTU) or genus-level
relative abundances the AUROC values were not significantly different
from the models trained without the SCFA concentrations
(P\textgreater{}0.21; Figure 2A). These data demonstrate that knowledge
of the SCFA profile from a patient's fecal sample does not improve the
ability to diagnose a colonic lesion.

\textbf{Knowledge of microbial community structure does not predict SCFA
concentrations.} We next asked whether the fecal community structure was
predictive of fecal SCFA concentrations, regardless of a person's
diagnosis. We trained Random Forest regression models using 16S rRNA
gene sequence data binned into OTUs and genera to predict the
concentration of the SCFAs (Figure S2). The largest R\textsuperscript{2}
between the observed SCFA concentrations and the modeled concentrations
was 0.14 observed when using genus data to predict butyrate
concentrations (Figure 2B). We also used a smaller dataset of shotgun
metagenomic sequencing data generated from a subset of our cohort
(N\textsubscript{normal}=27, N\textsubscript{adenoma}=25, and
N\textsubscript{cancer}=26) (18). We binned genes extracted from the
assembled metagenomes into operational protein families (OPFs) or KEGG
categories and trained Random Forest regression models using metagenomic
to predict the concentration of the SCFAs (Figure S2). Similar to the
analysis using 16S rRNA gene sequence data, the metagenomic data was not
predictive of SCFA concentration; the largest amount of variation that
the models could explain was 0.055, which was observed when using KEGG
data to predict propionate concentrations (Figure 2B). Because of the
limited number of samples that we were able to generate metagenomic
sequence data from, we used our 16S rRNA gene sequence data to impute
metagenomes that were binned into metabolic pathways or KEGG categories
using PICRUSt (Figure S2). SCFA concentrations could not be predicted
based on the imputed metagenomic data. The largest amount of variation
that the models could explain was 0.097 observed when using KEGG data to
predict butyrate concentrations (Figure 2B). The inability to model SCFA
concentrations from microbiome data indicates that the knowledge of the
abundance of organisms and their genes was insufficient to predict SCFA
concentrations.

\textbf{Conclusion.} Our data indicate that fecal SCFA concentrations
are not associated with the presence of adenomas or carcinomas and that
they provide weak predictive power to improve the ability to diagnose
someone with one of these lesions. Furthermore, knowledge of the
taxonomic and genetic structure of gut microbiota was not predictive
SCFA concentrations. These results complement existing literature that
suggest that fiber consumption and the production of SCFAs are unable to
prevent the risk of developing colonic tumors. It is important to note
that our analysis concerned fecal SCFA concentrations and microbiome
characterization and that observations along the mucosa near the site of
lesions may provide a stronger association. Regardless, given the
growing literature in this area, it is unlikely that SCFAs are the
primary mechanism that limits tumorigenesis. This may be a cautionary
result to temper enthusiasm for SCFAs as a biomarker of gut health more
generally. Going forward it is critical to develop additional hypotheses
for how the microbiome and host interact to drive tumorigenesis to
better understand the disease process and identify biomarkers that will
allow early detection of tumors.

@@

\includegraphics{../results/figures/classification_testing.pdf}

\includegraphics{../results/figures/regression_testing.pdf}
